---
title: "Homework Assignment 1"
author: |
  | Greg Forkutza
  | Student ID: 400277514
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    citation_package: natbib
    latex_engine: xelatex
    toc: true 
    toc_depth: 2  
fontsize: 10pt
linestretch: 1.5
geometry: margin=1in
urlcolor: blue
bibliography: A1.bib
header-includes:
  - "\\usepackage[nottoc]{tocbibind}"
---

\newpage

```{r setup, include=FALSE}
  
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(ggrepel)
require(knitr)
require(readr)
require(shiny)
require(rsconnect)

```


```{r, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, error=FALSE)
```


```{r, echo = FALSE, include = FALSE}

#  Download the zip file from the URL and convert to CSV
url <- "https://www150.statcan.gc.ca/n1/tbl/csv/13100416-eng.zip"
destfile <- "13100416-eng.zip"
download.file(url, destfile)

unzip(destfile)

data3 <- read_csv("13100416.csv")  %>%
         mutate(GEO = str_extract(GEO, "^[^,]+")) %>%
         mutate(GEO = case_when(
          GEO == "Newfoundland and Labrador" ~ "NL",
          GEO == "Prince Edward Island" ~ "PE",
          GEO == "Nova Scotia" ~ "NS",
          GEO == "New Brunswick" ~ "NB",
          GEO == "Quebec" ~ "QC",
          GEO == "Ontario" ~ "ON",
          GEO == "Manitoba" ~ "MB",
          GEO == "Saskatchewan" ~ "SK",
          GEO == "Alberta" ~ "AB",
          GEO == "British Columbia" ~ "BC",
          GEO == "Yukon" ~ "YT",
          GEO == "Northwest Territories" ~ "NT",
          GEO == "Nunavut" ~ "NU",
          TRUE ~ GEO
       )) %>%
        mutate(`Age of mother` = sub("Age of mother, ", "", `Age of mother`))
```

# Report

## Question(i)

The [third goal](https://sdgs.un.org/goals/goal3) of the UN sustainable 
development project is to ensure healthy lives and to promote well-being for all
at all ages. This includes ensuring that women are able to have healthy 
pregnancies and raise their children. Being pregnant at an age for which they
are better able to support themselves. On the other hand, if the number of
older pregnancies increases that may say something about a down turned economic 
situation in Canada. This could mean that the cost of living is too high and
therefore women have to wait much longer before they are financially
comfortable to have children. Lastly as the population of Canada increases,
we should expect the birth rate to increase. This may not be the case if the 
above is true. This might indicate also a downturn in the health and well being
of women in Canada. 


This data set is called
[Live births, by age of mother](https://open.canada.ca/data/en/dataset/5441dd55-0c1f-4774-94a5-f641f47eaf22/resource/5a990841-abd5-4967-97b9-515dbe1286f2)
It is sourced from the 
[Statistics Canada, Canadian Vital Statistics, Birth Database (CVSB)](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=3231)
This is an administrative survey that collects demographic information annually 
for any Canadian resident anywhere in the world. The birth is assigned to the 
mothers province of resident. 

The data covers the years 1991 to 2021. There is one categorical 
variable. This category is the age of the mother. There are two quantitative 
variables, the percentage (relative to that province and time period) 
and the number of live births in that age category at the time period.

By exploring size and distribution of the age of live births across Canada we 
can better understand how the health and well being of women and motherhood is 
changing over time.


## Question (ii)


The category `GEO` records the province of residence of the mother. If the 
province of residence is unknown it is given the level of 
`Unknown province or territory`. However there are no non zero values over the
entire time period so we can safely drop this level \cite{wickham2019}. 

For each province of the mother's residence and each time period there are two
numerical values; the number of live births and the percentage of live births 
within that time period over all levels of `Age of mother`. Since we want to 
better understand not only the distribution of the age of the mother but 
the change in number of births over time, we will drop the 
`Percentage of live births` value and keep only the
`Number of live births` value. The time series of the number of live births
plotted simultaneously across all levels will convey both the change in 
distribution and magnitude over time \cite{wickham2016}. 

Fortunately, because records of birth are non-voluntary, the data set does not
have any missing values. This is count data from birth records over levels
of a single category. 

There is probably not any outliers for the same reason as above. If there was 
some other sociological or economic information available to look at why 
unexpected changes in the gradient of the birth rate for that level occurred for 
a particular time period, then we might have reason to smooth these. However,
there is no reason with the information given to indicate the presence of
any outliers as the survey nature of the data is very high quality and
reliable. 

## Question (iii)

We can visualize across each province, the number of births
in each age group. However the number of teen pregnancies($age \leq 19$) is 
small relative
to the other age groups. This is the same with older pregnancies
($40 \leq age \leq 49$). Therefore
we plot these four age categories separately in order to be able to better 
visualize their trends. Using 
a log transformation was considered but it did not provide clarity of 
the trends across all age categories as it caused the slope of the curves 
to decrease towards horizontal and lose visual information. 


```{r, echo =FALSE, fig.width=10}
# remove "Age of mother, all ages" and  "Canada"
filtered_data2 <- data3 %>%
  filter(grepl("Number of live births", `Characteristics`))  %>%
  filter(GEO != "Unknown province or territory") %>%
  filter(GEO != "Northwest Territories including Nunavut") %>%
   filter(!(`Age of mother` %in% c("15 to 19 years", 
                                  "all ages", 
                                  "under 15 years",
                                  "not stated",
                                  "40 to 44 years",
                                  "45 to 49 years"
                                  ))) %>%
  group_by(`Age of mother`) 

# plot
ggplot(filtered_data2, aes(x = REF_DATE, y = VALUE, group = `Age of mother`, color = `Age of mother`)) +
  geom_line() +
  facet_wrap(~GEO, scales = "free", ncol = 4) +
  labs(title = "Number of Live Births by Age of Mother Across Provinces and Territories",
       x = "Year",
       y = "Number of Live Births") +
  theme_minimal() +
  scale_x_continuous(breaks = c(min(filtered_data2$REF_DATE), max(filtered_data2$REF_DATE)))

```


```{r, echo = FALSE, fig.width= 10}

filtered_data3 <- data3 %>%
  filter(grepl("Number of live births", `Characteristics`))  %>%
  filter(GEO != "Unknown province or territory") %>%
  filter(GEO != "Northwest Territories including Nunavut") %>%
  filter((`Age of mother` %in% c("under 15 years",
                                "15 to 19 years",
                                "40 to 44 years",
                                "45 to 49 years"
                                ))) %>%
  group_by(`Age of mother`)


ggplot(filtered_data3, aes(x = REF_DATE, y = VALUE, group = `Age of mother`, color = `Age of mother`)) +
  geom_line() +
  facet_wrap(~GEO, scales = "free", ncol = 4) +
  labs(title = "Number of Teenage and Older Pregnancies Across Provinces and Territories",
       x = "Year",
       y = "Number of Live Births") +
  theme_minimal()+
  scale_x_continuous(breaks = c(min(filtered_data3$REF_DATE), max(filtered_data3$REF_DATE)))
```

We see in the majority of provinces, the mean level of `Age of mother` goes from
25 to 29 years old to 30 to 34 years old. We observe a decrease in the number
of births for women less than 30 years old across all
provinces over the 31 year time period. Interestingly we see large
positive gradients in the curves for women 35 to 39 years old and 40 to 44 years
old in the provinces 
with the largest economies: Ontario, BC, AB, QC. We also notice that teen 
pregnancy decreases dramatically everywhere except in Nunavut. In conclusion we 
see that women on average are having babies at an older age across all 
provinces. This can be interpreted both positively and negatively: older mothers
can mean both that women are not pressured into motherhood early and that they 
lack the sufficient resources for motherhood early. We need to know the 
distribution of the age of mother as a function of economic and ethnic 
characteristics. Then, we can say whether the health and well being of mothers
and motherhood is increasing or decreasing and for whom specifically. 

## Question (iv)

```{r, echo=FALSE, fig.width= 10, fig.height=2}
filtered_data4 <- data3 %>%
  filter(grepl("Number of live births", `Characteristics`))  %>%
  filter(GEO == "Canada") %>%
  filter(`Age of mother` %in% "all ages") %>%
  group_by(`Age of mother`)


ggplot(filtered_data4, aes(x = REF_DATE, y = VALUE, group = `Age of mother`, color = `Age of mother`)) +
  geom_line() +
  facet_wrap(~GEO, scales = "free", ncol = 4) +
  labs(title = "Number of Live Births Aggregated Over All Age Groups in Canada ",
       x = "Year",
       y = "Number of Live Births") +
  theme_minimal()

```

We see that the number of live births, aggregating over all levels of age, 
provinces and territories is decreasing. This period of a sharp negative 
gradient begins in 2016. This is when many would remember the cost of living in
Canada to begun to have increased dramatically. So even as the population
increases in Canada we have a decrease in the birth rate. Again, demographic
data of the mothers and households would make this exploration more interesting.
This suggests a decrease in the health and well being of the average women in
Canada.

## Question (v)

The link to the shiny app is [HERE](https://forkman39.shinyapps.io/shinya1/).
This app allows you to view the time series of each level of the age of mother
for a particular province. The interpretation of these plots is the same
as in sections (iii) and (iv).



\newpage

# Supplementary Material 

```{r, eval =FALSE}
#  Download the zip file from the URL and convert to CSV
url <- "https://www150.statcan.gc.ca/n1/tbl/csv/13100416-eng.zip"
destfile <- "13100416-eng.zip"
download.file(url, destfile)

unzip(destfile)

# Change province columns to be more compact
data3 <- read_csv("13100416.csv")  %>%
  mutate(GEO = str_extract(GEO, "^[^,]+")) %>%
  mutate(GEO = case_when(
    GEO == "Newfoundland and Labrador" ~ "NL",
    GEO == "Prince Edward Island" ~ "PE",
    GEO == "Nova Scotia" ~ "NS",
    GEO == "New Brunswick" ~ "NB",
    GEO == "Quebec" ~ "QC",
    GEO == "Ontario" ~ "ON",
    GEO == "Manitoba" ~ "MB",
    GEO == "Saskatchewan" ~ "SK",
    GEO == "Alberta" ~ "AB",
    GEO == "British Columbia" ~ "BC",
    GEO == "Yukon" ~ "YT",
    GEO == "Northwest Territories" ~ "NT",
    GEO == "Nunavut" ~ "NU",
    TRUE ~ GEO
  )) %>%
  mutate(`Age of mother` = sub("Age of mother, ", "", `Age of mother`))

# remove "Age of mother, all ages" and  "Canada"
filtered_data2 <- data3 %>%
  filter(grepl("Number of live births", `Characteristics`))  %>%
  filter(GEO != "Unknown province or territory") %>%
  filter(GEO != "Northwest Territories including Nunavut") %>%
  filter(!(`Age of mother` %in% c("15 to 19 years", 
                                  "all ages", 
                                  "under 15 years",
                                  "not stated",
                                  "40 to 44 years",
                                  "45 to 49 years"
  ))) %>%
  group_by(`Age of mother`) 

# Plot
ggplot(filtered_data2, aes(x = REF_DATE, y = VALUE, group = `Age of mother`,
                           color = `Age of mother`)) +
  geom_line() +
  facet_wrap(~GEO, scales = "free", ncol = 4) +
  labs(title = "Number of Live Births by Age of Mother Across Provinces 
                and Territories",
       x = "Year",
       y = "Number of Live Births") +
  theme_minimal() +
  scale_x_continuous(breaks = c(min(filtered_data2$REF_DATE),
                                max(filtered_data2$REF_DATE)))

# Remove teen preganancies and older pregnancies 
filtered_data3 <- data3 %>%
  filter(grepl("Number of live births", `Characteristics`))  %>%
  filter(GEO != "Unknown province or territory") %>%
  filter(GEO != "Northwest Territories including Nunavut") %>%
  filter((`Age of mother` %in% c("under 15 years",
                                 "15 to 19 years",
                                 "40 to 44 years",
                                 "45 to 49 years"
  ))) %>%
  group_by(`Age of mother`)

# Plot
ggplot(filtered_data3, aes(x = REF_DATE, y = VALUE, group = `Age of mother`, 
                           color = `Age of mother`)) +
  geom_line() +
  facet_wrap(~GEO, scales = "free", ncol = 4) +
  labs(title = "Number of Teenage and Older Pregnancies 
                Across Provinces and Territories",
       x = "Year",
       y = "Number of Live Births") +
  theme_minimal()+
  scale_x_continuous(breaks = c(min(filtered_data3$REF_DATE),
                                max(filtered_data3$REF_DATE)))


# Include only Canada 
filtered_data4 <- data3 %>%
  filter(grepl("Number of live births", `Characteristics`))  %>%
  filter(GEO == "Canada") %>%
  filter(`Age of mother` %in% "all ages") %>%
  group_by(`Age of mother`)


# Plot
ggplot(filtered_data4, aes(x = REF_DATE, y = VALUE, group =
                             `Age of mother`, color = `Age of mother`)) +
  geom_line() +
  facet_wrap(~GEO, scales = "free", ncol = 4) +
  labs(title = "Number of Live Births Aggregated Over All Age 
                Groups in Canada ",
       x = "Year",
       y = "Number of Live Births") +
  theme_minimal()

```

\newpage


